\section{Disc-planet simulations with a dynamic viscosity
  prescription}
Consider the kinematic viscosity prescription 
\begin{align}\label{dyn-visc}
  \hat{\nu} = \hat{\nu}_0\left[1 + \left(A_\nu -
  1\right)\exp{\left(-\frac{\int_z^\infty\rho
        dz^\prime}{\int_{z_\nu}^\infty\rho_i dz^\prime}\right)}\right],  
\end{align}
where $z=z_\nu(R)$ is a chosen transition height. Initially, in the active
or viscous layer ($z\gg z_\nu$) the kinematic viscosity $\hat{\nu}\to
A_\nu\hat{\nu}_0$; while in the dead or low-viscosity layer ($z\ll
z_\nu$) the kinematic viscosity $\hat{\nu}\to\hat{\nu}_0$. Unlike the
static profiles used previously, Eq. \ref{dyn-visc} allows the
kinematic viscosity to dynamically respond to the density
field, i.e. $\hat{\nu}$ is time-dependent.  

However, implementing Eq. \ref{dyn-visc} in our numerical disc model
requires integration in the cylindrical vertical direction on a
spherical grid. To avoid this complication, we  
assume vertical hydrostatic equilibrium is approximately maintained 
throughout the simulation, so that 
\begin{align}
  \rho(R,\phi, z) \simeq
  \frac{\Sigma(R,\phi)}{\sqrt{2\pi}H}\exp{\left(-\frac{z^2}{2H^2}\right)},  
\end{align}
and similarly for $\rho_i$, where $\Sigma$ is the total column
density and the thin disc approximation has been applied ($z\ll R$). 
The above viscosity prescription becomes
\begin{align}\label{dyn-visc-simple}
   \hat{\nu} \simeq \hat{\nu}_0\left\{1 + \left(A_\nu -
  1\right)\exp{\left[-\frac{\rho}{\rho_i}\frac{\erfc(\zeta/\sqrt{2})}{\erfc(\zeta_\nu/\sqrt{2})}\right]}\right\},
%\notag\\
%  &\simeq \hat{\nu}_0\left\{1 + \left(A_\nu -
%  1\right)\exp{\left[-\frac{\rho}{\rho_i}\frac{\erfc(\zeta/\sqrt{2})}{\erfc(\zeta_\nu/\sqrt{2})}\right]}\right\}.
\end{align} 
where $\zeta = z/H$. Eq. \ref{dyn-visc-simple}, though much simplified
from  Eq. \ref{dyn-visc}, still allows the kinematic 
viscosity to vary with density: a
positive density perturbation ($\rho>\rho_i$) causes a reduction in
kinematic viscosity and vice versa; while the explicit dependence on
$\zeta$ causes viscosity to increase from the midplane to the disc
atmosphere.    

Here we describe one simulation using Eq. \ref{dyn-visc-simple}. 
