\documentclass[useAMS,usenatbib]{mn2e}
\usepackage{epsfig,rotate,graphicx}
\usepackage[fleqn]{amsmath}
\usepackage{subfigure}
\usepackage{lscape}
\usepackage{bm}
\usepackage{paralist}

\newcommand{\p}{\partial}
\newcommand{\mnras}{MNRAS}
\newcommand{\apj}{ApJ}
\newcommand{\na}{New A}
\newcommand{\aap}{A\&A}
\newcommand{\apjl}{ApJL}
\newcommand{\araa}{ARAA}
\newcommand{\dd}{\delta}
\newcommand{\adot}{\dot{a}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\gtrsim}{\;\raisebox{-.8ex}{$\buildrel{\textstyle>}\over\sim$}\;}
\newcommand{\lesssim}{\; \raisebox{-.8ex}{$\buildrel{\textstyle<}\over\sim$}\;}
\newcommand{\anrev}{{\it ARA\&A, }}
\newcommand{\apjs}{{\it ApJS, }}
\newcommand{\icar}{{\it Icarus, }}
\newcommand{\mnr}{{\it MNRAS, }}
\newcommand{\nat}{{\it Nature, }}
\newcommand{\sci}{{\it Sci, }}
\newcommand{\ana}{{\it A\&A, }}
\newcommand{\anas}{{\it A\&AS, }}
\newcommand{\aaps}{{\it A\&AS, }}
\newcommand{\anar}{{\it A\&AR, }}
\newcommand{\prd}{{\it Phys. Rev. D }}
\newcommand{\qjl}{{\it QJRAS, }}
\newcommand{\sbar}{\bar{\sigma}}
\newcommand{\avg}[1]{\langle #1 \rangle}
%% \newcommand{\max}{\mathrm{max}}
%% \newcommand{\min}{\mathrm{min}}
\newcommand{\aziavg}[1]{\langle #1 \rangle_\phi}
\newcommand{\init}[1]{#1^{(0)}}
\newcommand{\bu}{\bm{u}}
\newcommand{\bv}{\bm{v}}
\newcommand{\rin}{r_\mathrm{in}}
\newcommand{\rout}{r_\mathrm{out}}
\newcommand{\dampin}{r_\mathrm{d,in}}
\newcommand{\dampout}{r_\mathrm{d,out}}
\newcommand{\psid}{\psi_\mathrm{d}}
\newcommand{\psitr}{\psi_\mathrm{tr}}
\newcommand{\psimax}{\psi_\mathrm{max}}
\newcommand{\Rtr}{R_\mathrm{tr}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\mmax}{m_\mathrm{max}}
\newcommand{\trel}{t_\mathrm{rel}}
\newcommand{\zeus}{{\tt ZEUS-MP }}
\newcommand{\pluto}{{\tt PLUTO }}
\newcommand{\ii}{\mathrm{i}}
%\newcommand{\avg}[1]{\langle{#1}\rangle}
\DeclareMathOperator{\erf}{erf}




\title[Vortices in viscous discs]{Testing large-scale 
  vortex formation against viscous layers in three-dimensional discs}

%% \author[Lin]{ Min-Kai Lin
%%   \thanks{E-mail: mklin924@cita.utoronto.ca} \\ 
%% Canadian Institute for Theoretical Astrophysics,  
%% 60 St. George Street, Toronto, ON, M5S 3H8, Canada \\
%% }

\author[Lin and Umurhan]{ Min-Kai Lin$^1$
  \thanks{E-mail: mklin924@cita.utoronto.ca} and Orkan M. Umurhan $^2$ \thanks{E-mail:TBC} \\
  $^1$ Canadian Institute for Theoretical Astrophysics,
  60 St. George Street, Toronto, ON, M5S 3H8, Canada \\
  $^2$ TBC
}

\begin{document}

\maketitle
\begin{abstract}
  Vortex formation through the Rossby wave 
  instability (RWI) in protoplanetary discs has been invoked to play a
  role in planet formation theory, and has been suggested to explain
  the observation of large dust asymmetries in transitional discs.   
  However, whether or not vortex formation operates in layered
  accretion discs, i.e. models of protoplanetary discs including dead zones, 
  has not been verified. As a first step toward testing the robustness of
  vortex formation in 3D discs, we present customized
  non-linear hydrodynamical 
  simulations of global 3D protoplanetary discs with an imposed kinematic
  viscosity that is height-dependent. Two types of numerical 
  experiments are performed:
  \begin{inparaenum}[(i)]
  \item non-axisymmetric instability of artificial
    density bumps in viscous discs;\label{exp1}  
  \item vortex-formation at planetary gap edges in layered discs.%; and 
%  \item vortex-formation in discs with a radial viscosity transition.
  \end{inparaenum}\,
   Experiment (\ref{exp1}) shows that the linear instability is
   largely unaffected by viscosity and remains dynamical.   
   %However, this assumes that a steady-state viscous disc with a
   %density bump can be constructed in the first place.  
   The disc-planet simulations  
   also show the initial development of vortices at gap edges, but in
   layered discs the vortices are transient structures which disappear  
   well into the non-linear regime. We suggest that vortex
   formation via the RWI requires low effective viscosity throughout the 
   vertical extent of the disc. 
\end{abstract}

\begin{keywords}
planetary systems: formation --- planetary systems:
protoplanetary discs
\end{keywords}


\section{Introduction}\label{intro}

%\begin{itemize}
%\item transitional disk observations. ALMA. large-scale
%  structures. %planet interpretation 
%\item planet(estimal) formation %alread invoked in planetestimal
%                                %formation (barge sommeria 95)
%\end{itemize}

Recent observations have revealed a class of transition discs ---
circumstellar discs which are dust poor in its inner regions ---  
with non-axisymmetric dust distributions in its outer 
regions \citep{brown09,mayama12,marel13}. 
One interpretation of such a non-axisymmetric structure is the presence of
a large-scale disc vortex, which is known to act as a dust trap
\citep{barge95,inaba06,birnstiel13,ataiee13}.  
Because of its occurence adjacent to the inner dust
hole, i.e. a cavity edge, it has been suggested that such a vortex is
a result of the Rossby wave instability: a hydrodynamical instability
that can develop in radially structured discs.  
%2D results and relation to PPI
%linear stability analysis
%nonlinear simulations
%dust capture 

Modern work on the RWI began with 2D linear stability analysis
\citep{lovelace99,li00}. These studies show that a disc with radially localized
structure, such as a surface density enhancement over a
radial lengthscale of order the local disc scale-height, is unstable to
non-axisymmetric perturbations, which grow on dynamical (orbital)
timescales. Early 2D non-linear hydrodynamic simulations showed that
the RWI leads to multi-vortex formation, followed by vortex merging into
a single large vortex in quasi-steady state \citep{li01,inaba06}. 

While these studies consider disc models with artificial radial
structure, it has recently been estabilished that a natural site for
the RWI is the edge of gaps induced by disc-planet interaction 
\citep{koller03,li05,valborro07,li09,lyra09b,lin10,lin11a}. Indeed, this has
been the proposed explanation for the recently observed dust asymmetry
in the Oph IRS 48 transition disc system \citep{marel13}.      

%3D results
An important extension to the aforementioned theoretical studies is
the generalization of the RWI to 3D 
discs. Both nonlinear 3D hydrodynamic simulations 
\citep{meheut10,meheut12b,lin12b} and 3D linear calculations
\citep{umurhan10,meheut12,lin12,lin13} have been carried out. 
These studies reveal that the RWI is a 2D instability,
in that there is negligible difference between growth rates obtained
from 2D and 3D linear calculations. The associated density and
horizontal velocity perturbations have weak vertical dependence; and
vertical velocities are small. In nonlinear hydrodynamic simulations,
the vortices are columnar and extend throughout the vertical domain
\citep{richard13}. 

%relevance to layered disc models
The RWI therefore appears to be a global instability in the direction
perpendicular to the disc midplane. The vortical perturbation involves
the entire column of fluid. Thus conditions away
from the disc midplane may have important effects on vortex-formation
via the RWI. For example, \cite{lin13a} only found linear
instability for certain upper disc boundary conditions. This issue is
relevant to protoplanetary disc models including `dead zones'.   

%the question 
It is believed that accretion in protoplanetary discs is driven by
magnetohydrodynamic (MHD) turbulence as a result of the magneto-rotational instability
\citep[MRI,][]{balbus91}. However, it is not clear if the MRI operates
throughout the vertical extent of the disc, because the midplane of protoplanetary
disc is dense and cold \citep{armitage11}. As a result, \cite{gammie96}
proposed the layered disc model: accretion due to MHD turbulence
is small near the midplane (the dead zone), while MHD turbulence driven accretion
operaties near the disc surface (the active zone). The layered
accretion disc model has been subject to numerous studies 
\citep[e.g.,][]{fleming03,terquem08,oishi09,dzy10,kretke10,okuzumi11,flaig11,landry13}.    
If MRI-driven accretion can be modelled through an effective 
viscosity \citep{balbus99}, this corresponds to a low viscosity midplane and 
high viscosity atmosphere. It is therefore valid to ask how such a
vertical disc structure would affect the RWI. 

This problem is partly motivated by viscous disc-planet
simulations which show that the gap-edge RWI only develops when the viscosity is
sufficiently small \citep{valborro06,valborro07,edgar08}.%and edgar 'vertical structure of gaps'
What happens if the effective 
viscosity near the midplane is sufficiently low for the RWI to
develop, but is too high away from the midplane? Does the
instability, and therefore vortex formation, still take place? 

In this work we examine vortex formation through the RWI in
layered discs. As a first study to this problem, we take an
experimental approach through customized numerical hydrodynamic
simulations. We simulate global 3D protoplanetary discs with an imposed
kinematic viscosity that varies with height above the disc
midplane. %because most simulations use a NS viscosity. full mri
          %simulation too expensive. comparison to previous 
The purpose of this paper is to demonstrate, through selected
simulations, the potential importance of layered disc structures on
vortex formation. We defer a detailed parameter survey to a future
study. 

This paper is organised as follows. The accretion disc model is 
set up in \S\ref{model} and the numerical simulation method described
in \S\ref{sims}. Results are presented in \S\ref{density_bump} for 
viscous discs initialised with a density bump. These 
simulations employ a special setup such that the density bump is not
subject to axisymmetric viscous diffusion. This allows one to
focus on the effect of layered viscosity on the linear
non-axisymmetric instability. \S\ref{disc-planet} revisits vortex
formation at planetary gap edges, but in 3D layered discs, 
where it will be seen that vortex formation can be surpressed by
viscous layers. \S\ref{summary} concludes this work with a discussion
of important caveats of the present disc models.   

\input{model}
\input{sims}
\input{bump}
\input{planet}
%%\input{dead}

\input{summary}

%% \section*{Acknowledgments}
%rp nelson

\bibliographystyle{mn2e}
\bibliography{ref}

% \appendix
% \input{appendix}
%\input{appendix2}

\end{document}
